---
title: "PS 270 Final Project"
author: Taylor D'Andrea
description: "Incarceration Trends Across American Counties"
output:
  distill::distill_article:
    self_contained: false
---

### Project thoughts

I am interested in exploring data related to incarceration trends in the United States. Over the course of American history, Criminal Justice policies have changed drastically and incarceration has increasingly accelerated. In my analysis, I'd like to conduct an analysis of the "who" and "why" of modern incarceration trends. 

### The Proposal
Does county type, in conjunction with gender and racial composition, bolster incarceration rates across America counties? In this study, I plan to analyze the extent to which living in an American urban center, across gender and racial lines, enhances the incarceration rates in that particular place. I hypothesize that the more urban the county, the greater the incarceration rate will be. And I expect this increase will carry across all genders and races, though I do predict a disproportionately higher rate of incarceration for men over women, as well as for Black, Latinx and Native American individuals over Whites. Much research in the fields of Criminal Justice and Sociology have found patterns of increased incarceration for American cities, American ethnic centers, as opposed to suburban or rural areas of the country. The more economically diverse, ethnically diverse, opportunity diverse a place, the increased likelihood there is for conflict and the potential for incarceration. The data sample I will be using to conduct this analysis is comprised of American incarceration trends between 1983 and 2016. The data for this research came from the National Corrections Reporting Program (NCRP), the Deaths in Custody Reporting Program (DCRP), the Annual Survey of Jails (ASJ), the Census of Jails (COJ), the Census of State and Federal Adult Correctional Facilities (CAF), the FBI Uniform Crime Reporting Program (UCR), the Centers on Disease Control (CDC) - the National Vital Statistics System Data -, the U.S. Department of Agriculture, and the U.S. Census Bureau. My unit of analysis is the American prison system. The explanatory variables of interest are "urbanicity" which is rated on a scale of rural, small/mid, suburban, and urban. The co-explanatory variables will be gender (male or female) and race (White, Black, Latinx, or Native American). The outcome variable will be the incarceration rate (incarcerated individuals in a country per un-incarcerated individual) of American counties, which will be studied according to their determined urbanicity rating. The incarceration rate will also be studied for different combinations of gender and race in urban, versus small/mid, suburban and rural American communities. My hypothesis will be proven correct if urban centers have higher incarceration rates than small/mid, suburban, or rural counties and if those rates are disproportionately higher for men and Black, Latinx, and Native American members of those communities. My hypothesis would be disproven if the incarceration rate for urban centers is not different or less than that of urban counties in the United States. My hypothesis would also be disproven if gender and/or race have the opposite or no effect on the incarceration rates I have predicted. 

## The Data 
```{r}
library(tidyverse)
incarceration <- read_csv("data/incarceration_trends.csv")
incarceration
```

### Data Visualization (1)

Effect of American County Type on Prison Incarceration (by Gender) 

```{r}
incarceration <- incarceration |>
  drop_na(urbanicity, total_prison_pop, total_pop, female_prison_pop, male_prison_pop, female_pop_15to64, male_pop_15to64)
incarceration
```

```{r}
incarceration <- incarceration |>
  mutate(
    Female = (female_prison_pop / female_pop_15to64) * 100,
    Male = (male_prison_pop / male_pop_15to64) * 100,
    urbanicity = recode(urbanicity, 
                        "rural" = "Rural",
                        "small/mid" = "Small/Midsize",
                        "suburban" = "Suburban", 
                        "urban" = "Urban"))

gender_prison_plot <- incarceration |>
  group_by(urbanicity) |> 
  summarize(
    n = n(),
    Female = mean(Female, na.rm = TRUE),
    Male = mean(Male, na.rm = TRUE), 
    .groups = 'drop') |>
  mutate(urban_prop = n / sum(n)) |>
  pivot_longer(
    cols = c(Female, Male),
    names_to = "Gender",
    values_to = "urban_proportion") |>
  
  ggplot(aes(x = urbanicity,
             y = urban_proportion,
             fill = Gender)) +
  geom_bar(stat = "identity", 
           position = "dodge") + 
  theme_bw() +
  labs(title = "County Type & Prison Incarceration in the US (by Gender)", 
       x = "American County Types", 
       y = "Proportional Incarceration Rate of Civilians Age 15-64 (%)", 
       caption = "Source: Vera Institute of Justice, 2020") +
  scale_fill_manual(values = c("Female" = "red4", 
                               "Male" = "blue4"))
gender_prison_plot
```

This bar graph displays the incarceration rate of American civilians between the ages of 15 and 64 as a function of county type, filtered by gender, male or female. The county type of American counties was determined as being either: Rural, Small/Midsize, Suburban, or Urban. The incarceration rate is here presented as a percent, calculated by finding the proportions within each county type of both incarcerated males and females. The above bars represent what percent of which gender was incarcerated in each county type. 

One of the most distinct relationships determined by this graph is that men are incarcerated at much higher rates than women are at all levels of county type. There are not drastic incarceration rates for women in any county type. The male demographic seems to tell a significantly different story. Urban men are incarcerated at the highest rates, suburban men at the lowest. Rural men have the second highest incarceration rates, followed by small/midsize communities in third. For women, the order of highest rates of incarceration to lowest goes as follows: Rural, Small/Midsize, Urban, then Suburban. Overall, there seems to be a correlation between county type and incarceration; not necessarily that a more or less populated place has higher or lower incarceration rates, but that each county type on its own draws together unique circumstances that influence their incarceration rates. 

### Data Visualization (2)

Effect of American County Type on Prison Incarceration (by Race)

```{r}
incarceration <- incarceration |>
  filter(aapi_prison_pop > 0, aapi_pop_15to64 > 0, 
         black_prison_pop > 0, black_pop_15to64 > 0, 
         latinx_prison_pop > 0, latinx_pop_15to64 > 0, 
         native_prison_pop > 0, native_pop_15to64 > 0, 
         white_prison_pop > 0, white_pop_15to64 > 0) |>
  mutate(
    Asian = (aapi_prison_pop / aapi_pop_15to64) * 100,
    Black = (black_prison_pop / black_pop_15to64) * 100, 
    Latinx = (latinx_prison_pop / latinx_pop_15to64) * 100,
    Native = (native_prison_pop / native_pop_15to64) * 100,
    White = (white_prison_pop / white_pop_15to64) * 100)

race_prison_plot <- incarceration |>
  group_by(urbanicity) |> 
  summarize(
    n = n(),
    Asian = mean(Asian, na.rm = TRUE),
    Black = mean(Black, na.rm = TRUE), 
    Latinx = mean(Latinx, na.rm = TRUE), 
    Native = mean(Native, na.rm = TRUE),
    White = mean(White, na.rm = TRUE), 
    .groups = 'drop') |>
  mutate(urban_prop = n / sum(n)) |>
  pivot_longer(
    cols = c(Asian, Black, Latinx, Native, White),
    names_to = "Race",
    values_to = "urban_proportion") |>
  
  ggplot(mapping = aes(x = urbanicity,
                               y = urban_proportion,
                               fill = Race)) +
  geom_bar(stat = "identity", 
           position = "dodge") + 
  theme_bw() +
  labs(title = "County Type & Prison Incarceration in the US (by Race)", 
       x = "American County Types", 
       y = "Proportional Incarceration Rate of Civilians Age 15-64 (%)", 
       caption = "Source: Vera Institute of Justice, 2020") +
  scale_fill_manual(values = c("Asian" = "violet",
                               "Black" = "navy",
                               "Latinx" = "red4", 
                               "Native" = "green4",
                               "White" = "orange2"))
race_prison_plot
```

This bar graph displays the incarceration rate of American civilians between the ages of 15 and 64 as a function of county type, filtered by race, Asian, Black, Latinx, Native American, or White. Similarly to the first data visualization, the county type of American counties was determined as being either: Rural, Small/Midsize, Suburban, or Urban. The incarceration rate is here also presented as a percent, calculated by finding the proportions within each county type of incarcerated Asian, Black, Latinx, Native and White Americans. The bars above represent the percent at which each race was incarcerated in the different county types.

The most distinct finding in this graph is how Black individuals are incarcerated at much higher rates than any other race in all county types. In Rural areas, Black and Native American individuals are incarcerated as significantly higher rates than both their Asian, Latin and white counterparts. In Small/Midsize, Suburban and Urban areas, Black individuals are incarcerated at the highest rates, followed by Native Americans, Latinx, White, then Asian individuals. Overall, minority communities are incarcerated at disproportionate rates across all county types. Similarly to the first data visualization, identity, race in this case, seems to have a profound impact on rates of incarceration, which a wider trend than county type alone. However, also similarly to the first data visualization, the degree of urbanization in different American county types does seem to create unique circumstances within itself that determines the rates at which different races are incarcerated.  

### Preliminary Analysis of the Research Question 

Research Question: Does county type, in conjunction with gender and racial composition, bolster incarceration rates across America counties?

So far, this analysis appears to be drawing together the conclusion that identity is a greater factor on incarceration that the county type an American citizen is living in. In the Gender graph, men are consistently incarcerated more than women in every county type. In the Race graph, Black Americans are incarcerated at significantly higher rates compared to all other races in every county type. 

Despite this, each county type does appear to be drawing together special circumstances that impact incarceration rates. In Rural communities, the incarceration rates are highest for men and women compared to all other county types. Also in Rural communities, Black, Native American, White and Asian communities see their highest rates of incarceration compared to all other county types. In Small/Midsize communities, men have the third highest and women have the second highest rates of incarceration. The rate of incarceration for Black Americans in Small/Midsize communities is nearly the same as their incarceration rate in Rural communities. This is also where the Latinx community faces their highest incarceration rates. In Suburban communities, there are the lowest incarceration rates for both men and women, as well as for Black, Latinx, White and Asian communities compared to all other county types. Urban communities see the second highest rate of incarceration for men and the second lowest for women. Black individuals are incarcerated at the highest rates, followed by Native Americans, Latinx, White, then Asian Americans. 

All considered, though the broadest trends of incarceration are not entirely dependent on county type summatively, each county community type individually does seem to create certain circumstances that impacts the differences in their incarceration rates.


## The Final Report

### Introduction 
an introduction where you introduce the research question and hypothesis and briefly describe why it is interesting; describe the research question and main hypothesis; describe why it is important. (1-2 paragraphs) (2pts)

### The Data 
a data section that briefly describes the data source, describes how the key dependent and independent variables are measured (e.g., a survey, statistical model, or expert coding), and also produces a plot that summarizes the dependent variable; you should note if your research design is cross-sectional (most projects will be of this type) or one of the other designs we discussed (randomized experiment, before-and-after, differences-in-differences); 2-3 paragraphs + plot visualizing main outcome of interest. (3pts)

*include a confounding example with census division in this section that can be discussed throughout*

*name code chunks*

### The Results 
a results section that contains a scatterplot, barplot, or boxplot of the main relationship of interest and output for the main regression of interest; plot of main analysis + regression output (needs informative labels, names with a small number of digits presented; use kable function) + 2-3 paragraphs of description and interpretation of the plots and regression (including interpreting the main coefficient of interest and describing if it is statistically significant and if we should interpret it causally). This section could be longer if you choose to include additional analyses. (8pts)

### Conclusion 
a brief (one paragraph) concluding section that summarizes your results, assesses the extent to which you find support for your hypothesis, describes limitations of your analysis and threats to inference, and states how your analysis could be improved (e.g., improved data that would be useful to collect); 1 paragraph (i) summarizing results and assessing the extent to which you find support for your hypothesis; you should interpret (in plain English) the main coefficient of interest in your regression. You should also comment on the statistical significance of the estimated coefficient and whether or not you believe the coefficient to represent a causal effect; (ii) describing limitations of the analysis and threats to inference (missing data, confounding, etc), and stating how you could improve your analysis if you had more time/money. (2pts)
